Cartograms with Dynamic Features

### 2022-04-02 paper/GHRVis.tex
draft: review

\begin{abstract}

\end{abstract}

@@

>>>>

\begin{abstract}
Cartograms are representations of hybrid geographical and abstract data based on a value-by-area mapping technique combining statistical and geographical information. Based on Dorling cartogram, Demers cartogram uses squares instead of circles to represent regions, enabling the comparison of regions in a more intuitive way with a higher screen space utilization. One of the disadvantages of Dorling cartogram and its variants is that regions may be displaced from their original positions, reducing the legibility, readability, and geographical accuracy. To address that, we propose a novel layout algorithm that integrates dynamic topological features such as rivers into Demers cartograms. We evaluate the proposed layout algorithm through a user study with an application to an Electronic Health Records (EHR) dataset. The results demonstrate that the proposed layout algorithm is able to preserve the original layout of the cartogram, while improving the legibility, readability, and geographical accuracy of the cartogram. 
\end{abstract}

@@

\input{section/appendix}

### 2022-04-02 paper/section/appendix.tex
draft: review

>>>>

\section{Reproject Coordinate Reference System (CRS) of Shapefiles}
Shapefiles obtained use different CRS for projections. The river shapefiles use EPSG:4326 (WGS-84), whereas the CCG shapefile uses EPSG:27700 (OSGB36 - British National Grid). We first load all three river shapefiles into QGIS, followed by the CCG shapefile. QGIS automatically prompts to transform the latter into WGS-84, see \figref{fig:reproject}. The reprojection result is shown in \figref{fig:all_rivers}.

{
\begin{figure}[h!]
    \centering
    \includegraphics[width=\columnwidth]{figure/qgis/reproject.png}
    \caption{QGIS interface. A window prompt to transform the CCG shapefile's CRS into WGS-84.}
    \label{fig:reproject}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=\columnwidth]{figure/qgis/all_rivers.png}
    \caption{The reprojection result produced by \figref{fig:reproject}, which includes the River Trent (red), River Great Ouse (green), and River Thames (orange).}
    \label{fig:all_rivers}
\end{figure}
}

### 2022-04-02 paper/section/data.tex
draft: review

Obtaining the heterogenous data can be challenging, especially when an EHR dataset is involved \cite{wang2021EHRa}. The data acquisition involves multiple steps. The first step is to obtain both geospatial boundaries and EHR data that match one another. The second step is to pre-process the EHR data to remove empty and erroneous values. The final step is to transform the data into a format that is suitable for cartograms.

@@

Geospatial boundaries, or shapefiles, were obtained from the sources described below.
{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=0.6\columnwidth]{figure/ccg.png}
    \caption{A map of 135 CCGs in England as of 2020, obtained from the Open Geography portalx \cite{opengeographyportalxOpen}.}
    \label{fig:ccg}
\end{figure}
}

@@

\begin{lstlisting}[caption={The query that downloads the shapefile of River Thames from OpenStreetMap via the Overpass Turbo API.}, label={overpass},captionpos=b]
    relation(2263653);>>;
    out skel;
\end{lstlisting}

>>>>

Obtaining the heterogenous data can be challenging, especially when an EHR dataset is involved \cite{wang2021EHRa}. The first step is to obtain both geospatial boundaries and EHR data that match one another. The second step is to pre-process the EHR data to remove empty and erroneous values. The final step is to transform the data into a format that is suitable for cartograms.

@@

Geospatial boundaries, or shapefiles, were obtained from the sources described below.
{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=0.6\columnwidth]{figure/ccg.png}
    \caption{A map of 135 CCGs in England as of 2020, obtained from the Open Geography portalx \cite{opengeographyportalxOpen} with EPSG:27700 (OSGB36 - British National Grid) as the Coordinate Reference System (CRS).}
    \label{fig:ccg}
\end{figure}
}

@@

\begin{lstlisting}[caption={The query that downloads the shapefile of River Thames from OpenStreetMap via the Overpass Turbo API.}, label={overpass},captionpos=b]
    relation(2263653);>>;
    // River Great Ouse: 2798097
    // River Trent: 2863468
    out skel;
\end{lstlisting}

### 2022-04-02 paper/section/introduction.tex
draft: review

\new{Our contributions include:}

\begin{itemize}
    \item A new variant of Demers cartograms that incorporates dynamic topological features to improve readability and recognizability.
    \item A novel layout algorithm that preserves node positions relative to dynamic topological features such as rivers.
    \item A user study evaluation of the technique with an application to EHRs
\end{itemize}

\new{One of the major challenges involves is how to develop a layout algorithm that handles different shapes. In other words, the layout algorithm is novel because it handles different types of nodes: rectangular representing regions and polylines representing rivers. Another challenge we overcome in developing the algorithm is to revolve stalemate situations introduced by dynamic topological features while minimizing the geographical error.}

>>>>

Our contributions include:

\begin{itemize}
    \item A new variant of Demers cartograms that incorporates dynamic topological features to improve readability and recognizability,
    \item A novel layout algorithm that preserves node positions relative to dynamic topological features such as rivers,
    \item A user study evaluation of the technique with an application to EHRs.
\end{itemize}

One of the major challenges involved is how to develop a layout algorithm that handles different shapes. In other words, the layout algorithm is novel because it handles different types of nodes -- rectangular representing regions and polylines representing rivers. Another challenge we overcome in developing the algorithm is to resolve stalemate situations introduced by dynamic topological features while minimizing the geographical error.

### 2022-04-02 paper/section/main.tex
draft: review

{
    \begin{figure}[tb!]
        \centering
        \includegraphics[width=\columnwidth]{figure/flowchart.png}
        \caption{A flowchart illustration of our layout algorithm. See also \algoref{alg:UpdateNodePosition} for more detail.}
        \label{fig:flowchart}
    \end{figure}

    \begin{figure}[tb!]
        \centering
        \includegraphics[width=\columnwidth]{figure/flowchart2.png}
        \caption{An awesome flowchart illustration of our layout algorithm. See also \algoref{alg:UpdateNodePosition} for more detail.}
        \label{fig:flowchart2}
    \end{figure}
}

\algoref{alg:UpdateNodePosition} and \figref{fig:flowchart} provide an overview of the process. \textbf{Initialization:} We first load and render the CCG geospatial boundaries. For each CCG we compute the centroid and represent it using a small unit size square node with initial size, $ \nodeSize $. We start with small unit size squares because they do not overlap initially. \textbf{Node layout:} The basic algorithm is as follows: we first apply the Fast Node Overlap Removal (FNOR) algorithm that solves the Variable Placement with Separation Constraints (VPSC) problem \cite{dwyer2006fast} in order to remove overlaps. We chose FNOR over other node overlap removal algorithms because FNOR is able to provide spread minimization and node movement minimization while maintaining a good level of global shape preservation \cite{chen2020Node}. We gradually increase the node size by 1 unit at a time to ensure smooth transitions. An increase in $ \nodeSize $ can cause the nodes to overlap. During overlap removal, we compute node trajectories (See \algoref{alg:check river intersection}) and translate nodes to their new position. Nodes that cross a river are translated back to their previous position. This procedure ends when 1) no node overlap is present; 2) no nodes cross a river. We then increase $ \nodeSize $ by one pixel and repeat the algorithm until the max node size $ \nodeSizeMax $ is reached. The gradual size increase process provides stability to the layout and helps minimize geographical error.

@@

    \textbf{Local variables:} \\
    $ \node \gets $ a node is an object with the following property: \\
    \-\hspace{1em} $ \node.x, \node.y, $ or $ \node(x,y) \gets $ the x and y coordinates of $ \node $ \\
    \-\hspace{1em} $ \nodeCross \gets True $ if $ \node $ crosses a river \\
    \-\hspace{1em} $ \nodePrevious \gets $ the previous position of $ \node $ \\
    \-\hspace{1em} $ \nodeFNOR \gets $ the translated position of $ \node $ \\

    \begin{algorithmic}[1]
        \Procedure{UpdateNodePosition}{}
        \While{$ \nodeSize < \nodeSizeMax $}

            \While{$ \nodeListCross = True $}

@@

                        \State $ \node(x,y) \gets \nodeFNOR(x,y) $
                        \State $ \nodeCross \gets $ \Call{TestRiverIntersection}{$ \nodeFNORLine $}

                        \If{$ \nodeCross = True $}

                            \State $ \nodeListCross \gets True $
                            \State $ \nodeStalemate \gets \nodeStalemate + 1 $

                            \If{$ \nodeStalemate < \stalemateMax $}
                                \State \hskip-2em \Comment{Transalte back to previous position}
                                \State $ \node(x,y) \gets \nodePrevious(x,y) $

@@

                                \State \Call{ProcessStalemate}{$ \node, \nodeFNOR $}
                                \State $ \nodeStalemate \gets 0 $ \Comment{Reset counter}

@@

We use rivers as topological boundaries and prevent nodes from crossing them. The resolution of rivers can be adjusted by the user, as shown in \figref{fig:river resolution}, the number of edges is reduced from its original 10,170 to 65. We do not require a pixel-level river resolution as Demers cartograms do not represent geography at this level od detail. Instead a resolution that matches the node size is used. Enabling river simplification reduces the number of edge intersection tests that need to be performed. When a node's position changes, we test if the node's trajectory intersects any segment of a river. See \algoref{alg:check river intersection}. A bounding box intersection test between the edge defined by node translation and river edges can be performed to reduce the number of edge intersection tests required.

@@

    \begin{algorithmic}[1]
        \Procedure{TestRiverIntersection}{$ \nodeLineNV $}

@@

As the FNOR always attempts to produce an optimal node layout where node distribution and translation are minimized, a node's translation path might repeatedly intersect a river due to congestion, creating a stalemate situation, as shown in \figref{fig:stalemate}. If a node is translated between two positions, $ \node $ and $ \nodeFNOR $, for $ \stalemateMax $ iterations, we introduce a heuristic solution: constructing a corridor to break up the congestion. A corridor, $ c $, is a rectangle with a width of $ \CorridorWidth $ and a length of $ \CorridorLength $, formed by deriving two edges $ \EdgeParallelA $ and $ \EdgeParallelB $ such that $ \EdgeParallelA \parallel \EdgeParallelA \parallel \nodeLineNtNc $ (See \figref{fig:corridor}C and D). All enclosed nodes are then translated by $ \nodeVectorTC $ to alleviate the congestion (See \figref{fig:corridor}D).

{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=\columnwidth]{figure/stalemate.png}
    \caption{A stalemate situation is when a node's translation path $ \nodeVectorCT $ (in iteration 1) intersects a river for $ \stalemateMax $ times. The node is translated back to its previous position (iteration 2). A stalemate often occurs when the area is congested and the node is unable to translate to a position without intersecting a river.}
    \label{fig:stalemate}
\end{figure}
}

>>>>

{
    \begin{figure}[tb!]
        \centering
        \includegraphics[width=\columnwidth]{figure/flowchart.png}
        \caption{A flowchart illustration of our layout algorithm. See also \algoref{alg:UpdateNodePosition} for more detail.}
        \label{fig:flowchart}
    \end{figure}
}

\algoref{alg:UpdateNodePosition} and \figref{fig:flowchart} provide an overview of the process. \textbf{Initialization:} We first load and render the CCG geospatial boundaries. For each CCG we compute the centroid and represent it using a small unit size square node with initial size, $ \nodeSize $. We start with small unit size squares because they do not overlap initially. \textbf{Node layout:} The basic algorithm is as follows: we first apply the Fast Node Overlap Removal (FNOR) algorithm that solves the Variable Placement with Separation Constraints (VPSC) problem \cite{dwyer2006fast} in order to remove overlaps. We chose FNOR over other node overlap removal algorithms because FNOR is able to provide spread minimization and node movement minimization while maintaining a good level of global shape preservation \cite{chen2020Node}. We gradually increase the node size by 1 unit at a time to ensure smooth transitions. An increase in $ \nodeSize $ can cause the nodes to overlap. During overlap removal, we compute node trajectories (See \algoref{alg:check river intersection}) and translate nodes to their new position. Nodes that cross a river are translated back to their previous position. If a node oscillates across a river, we process this as a stalemate. This procedure ends when 1) no node overlap is present; 2) no nodes cross a river. We then increase $ \nodeSize $ by one pixel and repeat the algorithm until the max node size $ \nodeSizeMax $ is reached. The gradual size increase process provides stability to the layout and helps minimize geographical error.

@@

    \textbf{Local variables:} \\
    $ \node \gets $ a node is an object with the following properties: \\
    \-\hspace{1em} $ \node.x, \node.y, $ or $ \node(x,y) \gets $ the x and y coordinates of $ \node $ \\
    \-\hspace{1em} $ \nodeCross \gets $ the counter for river crosses for $ \node $ \\
    \-\hspace{1em} $ \nodePrevious \gets $ the previous position of $ \node $ \\
    \-\hspace{1em} $ \nodeFNOR \gets $ the translated position of $ \node $ \\

    \begin{algorithmic}[1]
        \Procedure{UpdateNodePosition}{}
        \While{$ \nodeSize < \nodeSizeMax $}
            \State $ \nodeListCross \gets True $
            \While{$ \nodeListCross = True $}

@@

                        \State $ \node(x,y) \gets \nodeFNOR(x,y) $

                        \If{ \Call{TestIntersection}{$ \nodeFNORLine $} $ = True $}
                            \State $ \nodeCross \gets \nodeCross ++ $
                            \State $ \nodeListCross \gets True $

                            \If{$ \nodeCross < \stalemateMax $}
                                \State \hskip-2em \Comment{Translate back to previous position}
                                \State $ \node(x,y) \gets \nodePrevious(x,y) $

@@

                                \State \Call{ProcessStalemate}{$ \node, \nodeFNOR $}
                                \State $ \nodeCross \gets 0 $ \Comment{Reset counter}

@@

We use rivers as topological boundaries and prevent nodes from crossing them. The resolution of rivers can be adjusted by the user, as shown in \figref{fig:river resolution}, the number of edges is reduced from its original 10,170 to 65. We do not require a pixel-level river resolution as Demers cartograms do not need to represent geography at this level od detail. Instead a resolution that matches the node size is used. Enabling river simplification reduces the number of edge intersection tests that need to be performed. When a node's position changes, we test if the node's trajectory intersects any segment of a river. See \algoref{alg:check river intersection}. A bounding box intersection test between the edge defined by node translation and river edges can be performed to reduce the number of edge intersection tests required.

@@

    \begin{algorithmic}[1]
        \Procedure{TestIntersection}{$ \nodeLineNV $}

@@

As the FNOR always attempts to produce an optimal node layout where node distribution and translation are minimized, a node's translation path might repeatedly intersect a river due to congestion, creating a stalemate situation, as shown in \figref{fig:stalemate}. If a node is translated between two positions, $ \node $ and $ \nodeFNOR $, for $ \stalemateMax $ iterations (a user-adjustable parameter), we introduce a heuristic solution: constructing a corridor to break up the congestion. A corridor, $ c $, is a rectangle with a width of $ \CorridorWidth $ and a length of $ \CorridorLength $, formed by deriving two edges $ \EdgeParallelA $ and $ \EdgeParallelB $ such that $ \EdgeParallelA \parallel \EdgeParallelA \parallel \nodeLineNtNc $ (See \figref{fig:corridor}C and D). All enclosed nodes are then translated by $ \nodeVectorTC $ to alleviate the congestion (See \figref{fig:corridor}D).

{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=\columnwidth]{figure/stalemate.png}
    \caption{A stalemate situation is when a node's translation path $ \nodeVectorCT $ (in iteration 1) intersects a river $ \stalemateMax $ times. The node is translated back to its previous position (iteration 2). A stalemate often occurs when the area is congested and the node is unable to translate to a new position without intersecting a river.}
    \label{fig:stalemate}
\end{figure}
}

### 2022-04-02 paper/section/related.tex
draft: review

\new{Our work extends the algorithm described by \citea{tong2018cartograms} which incorporates static topological features into Demers cartograms. Making topological features dynamic further improves the geographical accuracy and legibility.}

@@

One of the disadvantages of Dorling and Demers cartograms is the legibility. The layout algorithms may displace regions far from their original position and making the maps more difficult to interpret. Our work extends Demers cartograms. We introduce a new type of dynamic topological feature, a river, aiming to improve the interpretation, readability, and geographical accuracy of this class of cartograms.

>>>>

Our work extends the algorithm described by \citea{tong2018cartograms} which incorporates a static topological feature into Demers cartograms. We introduce multiple features (rivers) into the layout (as opposed to a single). Making topological features dynamic further improves the legibility and geographical accuracy.

The way we evaluate the cartograms is also different.  \citeauthor{tong2018cartograms} count river crossings to evaluate error. Here our focus is on readability, thus we include a preliminary user study.

@@

One of the disadvantages of Dorling and Demers cartograms is legibility. The layout algorithms may displace regions far from their original position and making the maps more difficult to interpret. Our work extends Demers cartograms. We introduce a new type of dynamic topological feature, a river, aiming to improve the interpretation, readability, and geographical accuracy of this class of cartograms.
